---
title: "Banco de la Republica data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Banco de la Republica data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Tasa Politica Monetaria (TPM)

The `pulso_tpm()` function returns Colombia's monetary policy rate set by
the Junta Directiva del Banco de la Republica.

```{r, eval=FALSE}
library(pulso)
tpm_2024 <- pulso_tpm(start = "2024-01-01", end = "2024-12-31")
head(tpm_2024)
```

Data is returned as a tibble with three columns: `fecha` (Date), `valor`
(numeric, percentage points), and `serie` (always `"tpm"`).

When the live Banrep SDMX API (`totoro.banrep.gov.co`) is available,
`pulso_tpm()` fetches real-time data. If the API is down, a bundled
snapshot extending to 2026-04-21 is used automatically.

```{r, eval=FALSE}
# Force fixture (no network required)
tpm_all <- pulso_tpm(use_fixture = TRUE)
range(tpm_all$fecha)
```

## Coming in v0.2.0

- `pulso_ibr()` for the Indicador Bancario de Referencia (IBR)
- `pulso_banrep()` for generic Banrep SDMX series
