---
title: "Getting started with pulso"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with pulso}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Loading GEIH microdata with pulso

`pulso` provides programmatic access to Colombia's Gran Encuesta
Integrada de Hogares (GEIH), the household labor force survey
published monthly by DANE (Departamento Administrativo Nacional
de Estadística).

## Quick start

```{r, eval=FALSE}
library(pulso)

df <- pulso_load(year = 2024, month = 6, module = "ocupados")
```

The result is a tibble with the survey microdata. By default, all
columns are returned with their original DANE codes (e.g., P6020,
P3271).

## Accessing variable metadata

Pass `metadata = TRUE` to get DANE codebook information attached
to the result:

```{r, eval=FALSE}
df <- pulso_load(year = 2024, month = 6, module = "ocupados",
                 metadata = TRUE)
```

You can describe individual columns:

```{r, eval=FALSE}
cat(pulso_describe_column(df, "p6020"))
```

Or list metadata for all columns:

```{r, eval=FALSE}
metadata_summary <- pulso_list_columns_metadata(df)
print(metadata_summary)
```

## What is GEIH?

GEIH is Colombia's primary labor market survey, conducted monthly
since 2007. It collects data on:

- Labor force participation (employed, unemployed, inactive)
- Wages and informal employment
- Demographic characteristics (age, sex, education)
- Household composition

Microdata is freely published by DANE in monthly zip files.
`pulso` automates the download, parsing, and harmonization across
the four GEIH design epochs (2007-2018, 2019-2023, 2024-present,
plus the historical ECH 2000-2006).

## Comparison with the Python package

`pulso` (R) mirrors the API of `pulso-co` (Python). For example:

```python
# Python
import pulso
df = pulso.load(year=2024, month=6, module="ocupados", metadata=True)
print(pulso.describe_column(df, "P6020"))
```

```{r, eval=FALSE}
# R
library(pulso)
df <- pulso_load(year = 2024, month = 6, module = "ocupados",
                 metadata = TRUE)
cat(pulso_describe_column(df, "p6020"))
```

Both packages share the same canonical data files (sources.json,
variable_map.json, dane_codebook.json) via the monorepo at
https://github.com/Stebandido77/pulso.

## Caching

Downloaded microdata is cached at
`tools::R_user_dir("pulso", "cache")` to avoid re-downloading.
Pass `cache = FALSE` to force re-download.

## Coverage and limitations

This is `pulso` v0.1.0. Currently supported:

- Single year/month/module loads via `pulso_load()`
- Metadata access via `pulso_describe_column()` and
  `pulso_list_columns_metadata()`
- Coverage: 2007 to present

Planned for v0.2.0+:

- `pulso_load_merged()` for multi-period loads
- Area filtering (urbano/rural)
- Full harmonization across epochs
- Validation status checks

See the GitHub issues for roadmap and known limitations.
